---
title: "Statistics for Bioinformatics and eScience - Handin 1"
author: "Group 1"
date: "11/27/2019"
output: pdf_document
---

# 1 Blood types genetics

### Ex 1.1

$\Omega$ = {I^A^, I^A^i, I^B^I^B^, I^B^i, I^A^I^B^, ii}

### Ex 1.2

P(ii) = 0.25  
P(I^A^i, ii) = 0.75

### Ex 1.3

P(C = I^A^I^B^) = 0.25

### Ex 1.4

Phenotypes: AB, A, B  
Distribution: AB = 0.25; A = 0.375; B = 0.375

### Ex 1.5

P(I^B^i, I^B^I^B^ | B = I^A^I^B^) = 0.375

## 1.1 Some simulation in R

### Ex 2.1

```{r echo = TRUE, eval = FALSE}
sample(c("ii", "iB", "iA", "BB", "AA", "AB"), size = 1, replace = TRUE)
```

### Ex 2.2

```{r echo = TRUE, eval = TRUE}
offspringG <- function(Parent1, Parent2){
  Allele1 <- sample(unlist(strsplit(Parent1, split = "")), replace = TRUE, size=1)
  Allele2 <- sample(unlist(strsplit(Parent2, split = "")), replace = TRUE, size=1)
  return(paste(Allele1, Allele2, sep = ""))
}


offspringP <- function(Genotype){
  Phenotype <- ""
  if (Genotype == "ii"){
    Phenotype <- "Type O"
  }
  if (Genotype == "AB" | Genotype == "BA"){
    Phenotype <- "Type AB"
  }
  if (Genotype == "iB" | Genotype == "Bi" | Genotype == "BB"){
    Phenotype <- "Type B"
  }
  if (Genotype == "iA" | Genotype == "Ai" | Genotype == "AA"){
    Phenotype <- "Type A"
  }
  return(Phenotype)
}
```

### Ex 2.3

```{r echo = TRUE, eval = TRUE}
offspingiAiA <- c()
for(i in 1:1000){
  offspingiAiA[i] <- offspringG("iA", "iA")
}
sum(offspingiAiA == "ii")/length(offspingiAiA)
sum(offspingiAiA != "AA")/length(offspingiAiA)
```

### Ex 2.4

```{r echo = TRUE, eval = TRUE}
secondGen <- c()

for(i in 1:1000){
  A <- offspringG("iA", "iB")
  B <- offspringG("AB", "AB")
  C <- offspringP(offspringG(A, B))
  secondGen[i] <- C
}

secondGen
```

# 2 A case study of DNA sequence

### Ex 3.1

$\Omega$ = {AA, AT, AC, AG, TA, TT, TC, TG, CA, CT, CC, CG, GA, GT, GC, GG}

### Ex 3.2

P(ACG) = (0.25 $\times$ 0.25 $\times$ 0.25) $\times$ 3 = 0.047

### Ex 3.3

$\Omega$ = {Ø, A, T, C, AA, AT, AC, AAA, ...}  
P(length 10 G^C^) = (0.75)^10^ = 0.05

### Ex 3.4

1: P(CG|C) = 0.5 $\times$ 0.1 = 0.05  
2: P(CG|C) = 0.3 $\times$ 0.4 = 0.12

## 2.1 Some simulations in R

### Ex 4.1

```{r echo = TRUE, eval = TRUE}
ACGsim <- function(reps){
  count = 0
  
  for(i in 1:reps){
    sqce_5 <- paste(sample(c("A", "T", "G", "C"), replace = TRUE, size = 5), collapse = "")
    
    if(grepl("ACG", sqce_5, fixed = TRUE)){
      count <- count + 1
    }
  }
  return(count/reps)
}
ACGsim(10000)
```

### Ex 4.2

```{r echo = TRUE, eval = TRUE}
Gsim <- function(reps, length = 10){
  sqces_with_G <- 0
  for(i in 1:reps){
    seq <- sample(c("A", "C", "T", "G"), replace =TRUE, size = length)
    if ("G" %in% seq){
      sqces_with_G <- sqces_with_G + 1
    }
  }
  sqces_with_no_G <- reps - sqces_with_G
  return(sqces_with_no_G/reps)
}

Gsim(10000)
```

### Ex 4.3

```{r echo = TRUE, eval = TRUE}
simG <- c()
pG <- c()
for(i in 1:20){
  simG[i] <- Gsim(10000, length = i)
  pG[i] <- dbinom(i, size = i, prob = 0.75)
}

plot(1:20, simG, col = "blue", pch = 0, xlab = "number of reps")
points(1:20, pG, col = "red", pch = 1)
legend("topright",legend = c("simulation", "probability"), 
       col = c("blue", "red"), pch = c(0, 1))
```

# 3 A case study of neuronal data

### Ex 5.1 and 5.2

```{r echo = TRUE, eval = TRUE}
isidata <- read.table("neuronspikes.txt", col.names = "isi")
hist(isidata$isi, breaks = 50, probability = TRUE)

for(i in seq(0.5, 1.5, 0.1)){
  curve(dexp((x), rate = i), from = 0, to = 5, col= (i*10)-3, add = TRUE)
}

legend("topright", legend=c(seq(0.5,1.5,0.1)), 
       col=c(2:length(seq(0.5,1.5,0.1))), lty = 1)
```

A $\lambda$ value of 1.5 fits the data the best.

### Ex 5.3

```{r echo = TRUE, eval = TRUE}
pexp(1, rate = 1.5)
pexp(1.5, rate = 1.5) - pexp(0.5, rate = 1.5)
```

# 4 Brain cell database

### Ex 6.1

The option na.strings converts empty strings to NA values.  

```{r echo = FALSE, eval = TRUE}
cells <- read.csv("cell_types.csv", na.strings = "")
```


### Ex 6.2

The donor species column indicates which values are humans and which are mice.  

```{r echo = TRUE, eval = TRUE}
sum(cells$donor__species == "Homo Sapiens") / length(cells[,1])
```

### Ex 6.3

```{r echo = TRUE, eval = TRUE}
hist(cells$ef__peak_t_ramp[cells$donor__species == "Homo Sapiens"], 
     proba = TRUE,  xlab = "Rank Spike Time (s)", 
     breaks = 50, main = "Rank Spike Time frequencies for Homo Sapiens")

hist(cells$ef__peak_t_ramp[cells$donor__species == "Mus musculus"], 
     proba = TRUE, xlab = "Rank Spike Time (s)", 
     breaks = 50, main = "Rank Spike Time frequencies for Mus musculus")
```

### Ex 6.4

```{r echo = TRUE, eval = TRUE}
first_plot = FALSE

for(i in 1:3){
  curve(dlnorm(x, sdlog = 0.6, meanlog = i), from = 0, to = 40, 
        add = first_plot, col = i+1, ylab = "log density function")
  first_plot = TRUE
}

legend("topright", legend = 1:3, col = 2:4, lty = 1)
```

### Ex 6.5

**meanlog = 2** is best, based on the fit of the corresponding density function in the plot.

### Ex 6.6

```{r echo = TRUE, eval = TRUE}
hist(cells$ef__peak_t_ramp[cells$donor__species == "Homo Sapiens"], 
     proba = TRUE,  xlab = "Rank Spike Time (s)", 
     breaks = 50, main = "Rank Spike Time frequencies for Homo Sapiens")

curve(dlnorm(x, sdlog = 0.6, meanlog = 2), 
      from = 0, to = 40, add = TRUE, 
      col = 3, ylab = "log density function")
```

Yes, the density fits the data.

### Ex 6.7

```{r echo = TRUE, eval = TRUE}
sum(cells$donor__species == "Homo Sapiens" & cells$donor__sex == "Male")
sum(cells$donor__species == "Homo Sapiens" & cells$donor__sex == "Female")
```
